Predictive State Geometry via Cantor Embeddings and Wasserstein Distance

### 2022-02-07 paper/manuscript.tex
final paper minus anonymized github link

Stationary processes, due to their time-translation invariance, inherit the
fractal temporality of sequence space. This can be easily visualized: Given a
length-$L$ sample $x_1\dots x_L$, and $n,k>0$, take a sliding window of pasts
and futures, $(x_{t-n+1}\dots x_t, x_{t+1}\dots x_k)$ for $t=n,\dots,L-k$.  For
each past-future pair, compute the truncated Cantor embeddings on the
\emph{reversed} past and (unreversed) future: $\left(C(x_t \dots
x_{t-n+1}),C(x_{t+1}\dots x_k)\right)$. The resulting pairs of real numbers can
be plotted as $(x,y)$-values on a scatter plot. The fractal that emerges
contains, in essence, all information necessary to understand a process'
temporal structures. See Fig. \ref{fig:cantor} for examples and guidance on how
to interpret the visualization.

@@

  \underline{function \CantorWasserstein} ($n,k,x_1\dots x_L$)\;
  \Input{Integers $n,k$ of past and future lengths} 
  \Input{Length-$L$ sequence $x_1\dots x_L$ of observations} 
  \Output{List $\UnqPasts$ of unique pasts}
  \Output{List of lists $\Cantors$ of Cantor-embedded futures} 
  \Output{Matrix $\Wass$ of Wasserstein distances} 
  $\UnqPasts \leftarrow \left[\right]$\; 
  $\Cantors \leftarrow \left[\right]$\; 
  \For{$t\leftarrow n$ \KwTo $L-k$} { 
    $\overleftarrow{\mathbf{x}} \leftarrow \left[x_{t-n+1},\dots, x_{t}\right]$\;
    $\overrightarrow{\mathbf{x}} \leftarrow \left[x_{t+1},\dots, x_{t+k}\right]$\;
      \vspace{-1.5em}
      \begin{flalign*}
        p &\leftarrow \sum_{\ell=1}^L
        \frac{2J(\overrightarrow{x}_\ell)}{(2|\mathcal{X}|-1)^\ell}\ ; &&
      \end{flalign*}

      \eIf{$\overleftarrow{\mathbf{x}} \in \UnqPasts$} { 
        $\Append\ \overleftarrow{\mathbf{x}}$ \KwTo $\UnqPasts$\; 
        $\Append\ [p]$ \KwTo $\Cantors$\; 
        } { 
        $j \leftarrow \Index\left(\overleftarrow{\mathbf{x}},\UnqPasts\right)$\;
        $\Append\ p$ \KwTo $\Cantors_j$\; 
        } 
  }
  $K \leftarrow \Length(\UnqPasts)$\; 
  $\Wass \leftarrow \Matrix(K,K)$\;
  \For{$k\leftarrow 1$ \KwTo $K$} { 
    \For{$\ell\leftarrow 1$ \KwTo $K$} {  
      $\Wass_{k \ell}
        \leftarrow
        \Wasserstein(\Cantors_k,\Cantors_\ell)$\;
        $\Wass_{\ell k} \leftarrow \Wass_{k \ell}$\; } } 
  \KwResult{$\UnqPasts$,$\Cantors$,$\Wass$}
  \caption{Steps for converting a sequence of categorical time-series data
        into a labeled collection of empirical distributions of Cantor-embedded futures,
        and a matrix of Wasserstein distances from said distributions.}
  \label{alg:cantorwass}
\end{algorithm}

@@

Despite this high order of structural complexity, the predictive state geometry
reconstructed from a sufficiently large sample of the Morse-Thue process
recovers the neighborhoods of $[0,1]$ that are relevant to the dynamics of the
original logistic map. Said differently, there is a correspondence between each
past $x_{-n+1}\dots x_{0}$ and a subset $V_{x_{-n+1}\dots x_{0}}$, such that
$V_{x_1\dots x_n}$ is the set of all points $y$ for which $x(f^{-t}(y)) =
x_{-t}$ for $0\leq t < n$. (Here, $f(y) = ry(1-y)$ and $x(y)$ is the encoding
$y\mapsto \{0,1\}$.) As it happens, pasts $x_{-n+1}\dots x_{0}$ whose predictive
states are close under the Wasserstein distance are also pasts for which the
sets $f(V_{x_{-n+1}\dots x_{0}})$ are close. That is, they correspond to
predictively similar ranges of the logistic map variable.

@@

Similarly, there are many ways to metrize the product topology on sequences,
but the Cantor embedding offers a direct way to connect the product topology
with a visualizable geometry. And, embedding in a single dimension enables
efficient computation of the Wasserstein metric. The benefits of the Cantor and
Wasserstein approaches adds interpretability to the resulting predictive-state
geometry along two distinct axes, most clearly seen in Fig. \ref{fig:hclust}'s
clustered Cantor diagrams.

@@

\section{Reproducibility statement}
For the purposes of reproducibility, we provide a GitHub repository containing
the code necessary to generate this manuscript and its figures, including a
notebook for generating the data used for the examples and for building both
static and interactive figures for further exploration.
%%
%% The next two lines define the bibliography style to be used, and % the
%bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{ref}

>>>>

Stationary processes, due to their time-translation invariance, inherit the
fractal temporality of sequence space. This can be easily visualized: Given a
length-$L$ sample $x_1\dots x_L$, and $n,k>0$, take a sliding window of pasts
and futures, $(x_{t-k+1}\dots x_t, x_{t+1}\dots x_n)$ for $t=k,\dots,L-n$.  For
each past-future pair, compute the truncated Cantor embeddings on the
\emph{reversed} past and (unreversed) future: $\left(C(x_t \dots
x_{t-k+1}),C(x_{t+1}\dots x_n)\right)$. The resulting pairs of real numbers can
be plotted as $(x,y)$-values on a scatter plot. The fractal that emerges
contains, in essence, all information necessary to understand a process'
temporal structures. See Fig. \ref{fig:cantor} for examples and guidance on how
to interpret the visualization.

@@

  \underline{function \CantorWasserstein} ($k,n,x_1\dots x_L$)\;
  \Input{Integers $k,n$ of past and future lengths} 
  \Input{Length-$L$ sequence $x_1\dots x_L$ of observations} 
  \Output{List $\UnqPasts$ of unique pasts}
  \Output{List of lists $\Cantors$ of Cantor-embedded futures} 
  \Output{Matrix $\Wass$ of Wasserstein distances} 
  $\UnqPasts \leftarrow \left[\right]$\; 
  $\Cantors \leftarrow \left[\right]$\; 
  \For{$t\leftarrow n$ \KwTo $L-k$} { 
    $\overleftarrow{\mathbf{x}} \leftarrow \left[x_{t-k+1},\dots, x_{t}\right]$\;
    $\overrightarrow{\mathbf{x}} \leftarrow \left[x_{t+1},\dots, x_{t+n}\right]$\;
      \vspace{-1.5em}
      \begin{flalign*}
        p &\leftarrow \sum_{\ell=1}^n
        \frac{2J(\overrightarrow{x}_\ell)}{(2|\mathcal{X}|-1)^\ell}\ ; &&
      \end{flalign*}

      \eIf{$\overleftarrow{\mathbf{x}} \in \UnqPasts$} { 
        $\Append\ \overleftarrow{\mathbf{x}}$ \KwTo $\UnqPasts$\; 
        $\Append\ [p]$ \KwTo $\Cantors$\; 
        } { 
        $j \leftarrow \Index\left(\overleftarrow{\mathbf{x}},\UnqPasts\right)$\;
        $\Append\ p$ \KwTo $\Cantors_j$\; 
        } 
  }
  $K \leftarrow \Length(\UnqPasts)$\; 
  $\Wass \leftarrow \Matrix(K,K)$\;
  \For{$i\leftarrow 1$ \KwTo $K$} { 
    \For{$j\leftarrow 1$ \KwTo $K$} {  
      $\Wass_{ij}
        \leftarrow
        \Wasserstein(\Cantors_i,\Cantors_j)$\;
        $\Wass_{j i} \leftarrow \Wass_{i j}$\; } } 
  \KwResult{$\UnqPasts$,$\Cantors$,$\Wass$}
  \caption{Steps for converting a sequence of categorical time-series data
        into a labeled collection of empirical distributions of Cantor-embedded futures,
        and a matrix of Wasserstein distances from said distributions.}
  \label{alg:cantorwass}
\end{algorithm}

@@

Despite this high order of structural complexity, the predictive state geometry
reconstructed from a sufficiently large sample of the Morse-Thue process
recovers the neighborhoods of $[0,1]$ that are relevant to the dynamics of the
original logistic map. Said differently, there is a correspondence between each
past $x_{-k+1}\dots x_{0}$ and a subset $V_{x_{-k+1}\dots x_{0}}$, such that
$V_{x_{-k+1}\dots x_{0}}$ is the set of all points $y$ for which $x(f^{-t}(y)) =
x_{-t}$ for $0\leq t < n$. (Here, $f(y) = ry(1-y)$ and $x(y)$ is the encoding
$y\mapsto \{0,1\}$.) As it happens, pasts $x_{-k+1}\dots x_{0}$ whose predictive
states are close under the Wasserstein distance are also pasts for which the
sets $f(V_{x_{-k+1}\dots x_{0}})$ are close. That is, they correspond to
predictively similar ranges of the logistic map variable.

@@

Similarly, there are many ways to metrize the product topology on sequences, but
the Cantor embedding offers a direct way to connect the product topology with a
visualizable geometry. And, embedding in a single dimension enables efficient
computation of the Wasserstein metric. The benefits of the Cantor and
Wasserstein approaches adds interpretability to the resulting predictive-state
geometry along two distinct axes, most clearly seen in Fig. \ref{fig:hclust}'s
clustered Cantor diagrams. We hope that the success of this approach in
providing clear insights will complement existing thrusts in the direction of
abstract embeddings and mathematical formalism by motivating further development
on interpretable approaches to predictive state analysis.

@@

\section{Reproducibility statement}
For the purposes of reproducibility, we provide a 
GitHub repository \cite{repo} containing
the code necessary to generate this manuscript and its figures, including a
notebook for generating the data used for the examples and for building both
static and interactive figures for further exploration.
%%
%% The next two lines define the bibliography style to be used, and % the
%bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{ref}
